---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and transforming the data:
```{r load}
data <- read.csv("activity.csv", header = TRUE, sep = ",", na.strings = "NA")
summary(data)
str(data)
head(data)
data$date <- as.Date(data$date, format = "%Y-%m-%d")
data$interval <- factor(data$interval)
```

## What is mean total number of steps taken per day?

Data without NA values
```{r No-NA}
NA_index <- is.na(as.character(data$steps))
data_no_NA <- data[!NA_index,]
head(data_no_NA)
```

Calculating the total number of steps taken per day 
```{r totalSteps}
steps_each_day <- aggregate(steps ~ date, data = data_no_NA, sum)
colnames(steps_each_day) <- c("date", "steps")
```

Histogram of the total number of steps taken per day
```{r hist}
hist(as.numeric(steps_each_day$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Histogram of the total number of steps taken each day")
```

Calculating the mean and median of the total number of steps taken per day
```{r mean}
#Mean
mean(steps_each_day$steps)
```

```{r median}
#Median
median(steps_each_day$steps)
```

## What is the average daily activity pattern?

Making a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r timeseriesplot}
steps_per_interval <- aggregate(data_no_NA$steps, by=list(interval=data_no_NA$interval), FUN=mean)
colnames(steps_per_interval) <- c("interval", "average_steps")
plot(as.integer(levels(steps_per_interval$interval)), steps_per_interval$average_steps, type="l", xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern",  col ="blue")
```

The five minute interval which has the maximum number of steps is
```{r maxsteps}
max_steps <- max(steps_per_interval$average_steps)
max_steps
interval_max_steps<-steps_per_interval[which.max(steps_per_interval$average_steps),]$interval
interval_max_steps
```
The 5-minute interval, on average across all the days in the data set is 835, and contains the maximum number of 206 steps.

## Imputing missing values

Total number of missing values in the data set
```{r n_NA}
sum(is.na(as.character(data$steps)))
sum(is.na(as.character(data$date)))
sum(is.na(as.character(data$interval)))
```

Filling the missing values in the data set
```{r fillMiss}
#finding the indices of missing values
NA_index <- which(is.na(as.character(data$steps)))
complete_data <- data
#Imputing missing values using the mean for that 5-minute interval
complete_data[NA_index, ]$steps<-unlist(lapply(NA_index,FUN=function(NA_index){steps_per_interval[data[NA_index,]$interval==steps_per_interval$interval,]$average_steps}))
```

Creating a new data set with the missing values filled in
```{r imputedDataFrame}
#Creating a data frame with the steps taken for each day
steps_each_day_complete <- aggregate(steps ~ date, data = complete_data, sum)
#Adding column names to the created data frame
colnames(steps_each_day_complete) <- c("date", "steps")
```


Histogram of the total number of steps taken per day
```{r histogram}
hist(as.numeric(steps_each_day_complete$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Histogram of the total number of steps taken each day")
```

Calculating the mean and median of the total number of steps taken per day
```{r mean_complete}
#Mean
mean(steps_each_day_complete$steps)
```

```{r median_complete}
#Median
median(steps_each_day_complete$steps)
```
It can be seen that the mean is same as before and the median does not differ much before and after imputing

## Are there differences in activity patterns between weekdays and weekends?

Created a plot to compare and contrast number of steps between the week and weekend. There is a higher peak earlier on weekdays, and more overall activity on weekends.
```{r plot}
complete_data$day <- as.factor(weekdays(complete_data$date))
complete_data$is_weekday <- ifelse(!(complete_data$day %in% c("Saturday","Sunday")), TRUE, FALSE) 
weekdays_data <- complete_data[complete_data$is_weekday,]
steps_per_interval_weekdays <- aggregate(weekdays_data$steps, by=list(interval=weekdays_data$interval), FUN=mean)
weekends_data <- complete_data[!complete_data$is_weekday,]
steps_per_interval_weekends <- aggregate(weekends_data$steps, by=list(interval=weekends_data$interval), FUN=mean)
colnames(steps_per_interval_weekdays) <- c("interval", "average_steps")
colnames(steps_per_interval_weekends) <- c("interval", "average_steps")
steps_per_interval_weekdays$day <- "Weekday"
steps_per_interval_weekends$day <- "Weekend"
week_data <- rbind(steps_per_interval_weekends, steps_per_interval_weekdays)
week_data$day <- as.factor(week_data$day)
library(lattice)
xyplot(average_steps ~  interval | day, data = week_data, layout = c(1,2), type ="l", ylab="Number of Steps")
```




